\name: A Non-Parametric, Hierarchical Wavelength-Calibration Model for a Precision Spectrograph

### 2020-05-01 paper/paper.tex
Changes up to section 2.1

\documentclass[twocolumn]{aastex63}
\usepackage{xcolor}
\newcommand{\lz}[1]{\textcolor{orange}{#1}}

@@

% typesetting words
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\name}{\project{Excalibur}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\expres}{\project{\acronym{EXPRES}}}
\newcommand{\eprv}{\acronym{EPRV}}
\newcommand{\lfc}{\acronym{LFC}}

@@

\begin{document}
\title{\name:
A non-parametric, hierarchical model for a precision spectrograph}

@@

\begin{abstract}
\name\ is a hierarchical, non-parametric framework for wavelength calibrating precision spectrographs.  This method makes use of the many, recent hardware improvements in highly-stabilized, extreme-precision radial-velocity spectrographs.  With stabilized spectrographs, the full optical system and detectors can only vary within a tiny range of configurations.  This allows us to use all calibration data to determine the entire space of possible calibrations for a stabilized spectrograph.  A wavelength solution then need only pinpoint where in this low-dimensional calibration space the spectrograph exists.  \name\ also takes advantage of the development of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density of points permits the freedom of a non-parametric wavelength solution that can match or adapt to any instrument or detector oddities.  We demonstrate the success of this method with data from a Menlo Systems laser frequency comb taken by \expres, the Extreme Precision Spectrograph.  We find that \name\ returns more accurate wavelength predictions by more than a factor of 2 over classic polynomial models.  Employing \name\ wavelengths reduced the RMS of RV measurements to a planet fit by \lz{some percent}.
\end{abstract}

% Add official keywords
\keywords{instrumentation: spectrographs -- techniques: spectroscopic -- techniques: radial velocities}

\section{Introduction} 
Precision radial-velocity (\eprv) programs have been incredibly fruitful in finding and characterizing extra-solar planets \citep{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $10^5$, which correspond to line widths on the order of $3000\,\mps$.  Exoplanet science happens at the level of $3\,\mps$ precision, with current instruments hoping to reach $0.1\,\mps$ precision \citep{fischer2016}.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrograph is well sampled).    Some hardware--software systems, such as ESPRESSO and \expres\ are now reaching these levels \citep{pepe2013,  jurgenson2016, neid, blackman2020}.

Here we propose to simplify and improve calibration programs for \eprv\ hardware systems with two very simple but novel ideas.  The first flows from the observation that calibration sources---which include lamps \lz{(looks like the ThAr isn't quite good enough actually)}, etalons, and laser-frequency combs (\lfc s)---illuminate the spectrograph with very stable, very dense sets of lines; almost every location in the spectrograph image plane is surrounded by nearby, useful calibration lines.  This recommends a calibration methodology that is \emph{non-parametric}:  If every point in the spectrograph detector is surrounded by nearby calibration lines, the wavelength solution can, for example, be made simply as an interpolation of the calibration data.

The density of lines removes the need to enforce any functional form for the wavelength solution (such as a continuous eighth-order polynomial, for example).  We expand on the work of \cite{milakovic2020}, which demonstrated the efficacy of constructing a wavelength solution as multiple, segmented polynomials.  Going non-parametric will improve calibration accuracy by not forcing the choice of a parametric form that may bias the calibration, especially when the chosen function is inappropriate (as, for example, polynomials will be at detector edges).

The second idea follows from the observation that contemporary \eprv\ instruments are incredibly stable.  Temperature-controlled, fiber-fed spectrographs vary only slightly over the night or season, and only along a small number of axes in what you might call ``calibration space'', or the (very high dimensional) space of all possible wavelength solutions.  That is, not only are the spectrographs stable, but they also have few accessible degrees of freedom.  This renders it inappropriate to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph existed for each exposure.

This structure is \emph{hierarchical}: the calibration data are used not just to determine the wavelength solution, but also to determine the possible space of wavelength solutions.  In the statistics literature this concept is often described as \emph{de-noising}.  Every calibration exposure contains information about every other exposure.  Thus every exposure can be improved (i.e. de-noised) with information from every other exposure.

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven model for the wavelength solution.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the observed variations in the stabilized spectrograph.  \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, etalons, or good arc lamps.  We have in mind the \eprv\ instruments and \eprv\ science cases, but we expect \name\ to have applications for other kinds of spectrographs in other contexts.  The motivating ideas behind this project are true for almost every astronomical spectrograph.

\section{Method} \label{sec:method}
\name\ is designed to turn a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength solution.  We operate on the two core ideas that the wavelength solution should be given enormous flexibility, but that it lives in a very low-dimensional space, where the degrees of freedom are set by the limited kinematics of the stabilized, spectrograph hardware.

We therefore assume dense enough calibration line coverage with well-fit line centers to provide some constraint on an interpolated wavelength solution across an echelle order.  Additionally, we assume the instrument hardware is stabilized enough that the space of possible calibration states is low-dimensional.

We will pose the question of wavelength calibration in the following way.  Given an exposure $n$, and echelle order $m$, there is a relationship between
the two-dimensional $(x,y)$-position on the detector and the
wavelength $\lambda$
\begin{equation}
\lambda(x,y,m,n) = f(x,y,m;\theta_{n})
\quad ,
\label{eq:wsol}
\end{equation}
where $\theta_{n}$ represents the parameters describing a given exposure.

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the pipeline described in \cite{petersburg2020} sets the function $f(x,y,m;\theta_{n})$ from Equation \ref{eq:wsol} to be a 2D, 9\textsuperscript{th}-order polynomial where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$.
\begin{equation}
\lambda(x,y,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
The coefficients $c_{nij}$ are then smoothed by a third-order polynomial over time.  This third-order polynomial is evaluated at the time of non-calibration exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image is fit to a 2D polynomial independently.

@@

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that we can linearize, then the function $f(x,y,m;\theta_{n})$ from Equation \ref{eq:wsol} could be replaced with a tiny model
\begin{equation}
\lambda(x,y,m,n) = g_0(x,y,m) + \sum_{k=1}^K a_{nk}\,g_k(x,y,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(x,y,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are scalar amplitudes,
and the $g_k(x,y,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.

The challenge is to learn these basis functions from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern these basis functions.  In this paper, we present a model using principal component analysis (PCA).  This is the correct thing to do in the limit where we have very high SNR, as is undeniably the case with calibration images.

\lz{Is this the place to mention options other than PCA? Or should we save that for the discussion?}

\subsection{De-Noising of Calibration Frames} \label{sec:denoising}
Within a given range of stability, we can use all calibration images to  1) determine the space in which the instrument varies and 2) where in the accessible calibration space the spectrograph existed for each exposure.  Say we have $N$ calibration exposures, $exp_n$.  The first task is to identify the full list of lines we can expect to find in a calibration image.  A line, $l_{m,\lambda}$ is uniquely defined by a combination of echelle order, $m$, and ``true" or theoretical wavelength, $\lambda$.

For each exposure, $exp_n$, every line, $l_{m,\lambda}$, has an associated fitted detector position, $x_{n,l}$, for example x-pixel in an 2D extracted echelle order.  Fitted line centers that are missing from an exposure (e.g. because the fit failed due to noise, the line is not in its usual echelle order, etc.) can be assigned a $NaN$ for that echelle order instead.  Let there be $M$ lines per exposure.  We can therefore construct a $N \times M$ matrix of line positions for each of $M$ lines for each of $N$ exposures.

The average line position in every exposure represents the fiducial, or standard calibration of the spectrograph, $g_0(x,y,m)$.  We apply principal component analysis to the difference between this fiducial calibration and each individual line position.  The returned principal components serve as basis functions,  $g_k(x,y,m)$, expressing the possible deviations of the spectrograph from this fiducial calibration.  The magnitude of each principal component for each exposure, $a_{nk}$, represents the scalar amplitude of these deviations for each exposure.  We can now use a small number, $K$, of principal components to reconstruct a de-noised version of the line positions as formulated in Equation \ref{eq:excl_wsol}.

Missing line center measurements are replaced with de-noised estimates.  We do this iteratively until the estimates of missing line centers change by less than 0.01\%.  This process can be repeated on line centers deemed as outliers by some metric, to account for lines that may have been mis-identified or mis-fit.  The principal components from the final iteration are used to define the spectrograph's calibration space while the associated magnitudes pinpoint where in that calibration space the spectrograph is for each calibration exposure.

\begin{algorithm}
\SetAlgoLined
\textbf{Inputs}: $m$, echelle order of each line; $\lambda$, assigned wavelength of each line; $[x_{n,l}] $, matrix of pixel line position of each line $l$ and for each exposure $n$\;
\While{change in missing or outlier line centers $>$ 0.01\%}{
	$g_0(x,y,m) = \overline{x_{n,l}]}$\;
	find $U, \Sigma, V$ s.t. $U\Sigma V^* = (x_{l,n}-g_0(x,y,m))$\;
	let $a_{n,k} = U\cdot \Sigma$ and $g_k(x,y,m) = V$\;
	$\lambda(x,y,m,n) = g_0(x,y,m) + \sum_{k=1}^K a_{nk}\,g_k(x,y,m)$ for $K=6$\;
	$[x_{n,l}] = \lambda(x,y,m,n)$ where $[x_{n,l}]$ was initially $NaN$
	}
\caption{De-Noising}
\end{algorithm}

\subsection{?Interpolating Calibration Frames?}
 \label{sec:interp_time}
The magnitude of each principal component can be interpolated with respect to housekeeping data, such as time, temperature, telescope position, etc. to determine different calibration states of the spectrograph.  The choice of what to interpolate with respect to depends on the dominant contribution to variation in the instrument.

As an example, we can interpolate the magntiudes of the principal components with respect to time to the midpoint time of non-calibration exposures to determine the calibration state of the spectrograph for that exposure.  Using interpolated magnitudes, $a'_{nk}$ and the basis vectors, $g_k(x,y,m)$, returned by the de-noising process, we can construct a set of calibration lines, $\lambda(x,y,m,n)$, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.

\subsection{Interpolating a Wavelength Solution} \label{sec:interp_wsol}
We now have the power to construct a set of calibration lines, $\{l_{m,\lambda}\}$, each with an assigned, known wavelength for any exposure.  To construct a wavelength solution, we need only interpolate the known wavelengths over the exposure specific line centers.  For instance, interpolating the known wavelengths vs. line centers onto every integer x will generate wavelengths for each pixel in an echelle order.

\begin{algorithm}
\SetAlgoLined
\textbf{Inputs}: $g_0(x,y,m)$, the fiducial calibration of the spectrograph; $a_{n,k}$, magnitudes of the principal components for each exposure; $g_k(x,y,m)$, basis vectors spanning the calibration space of the spectrograph; $t_n$, time of each exposure; $t'$, time we want wavelengths for\;
Interpolate $a_{n,k}$ with respect to time to determine $a_{n,k}'$ for time $t'$\;
Calculate pixel line positions for time $t'$ by evaluating\;
$\lambda(x,y,m,n)' = g_0(x,y,m) + \sum_{k=1}^K a_{nk}'\,g_k(x,y,m)$ for $K=6$\;
\For{each unique $m$}{
	interpolate ${l_n,m}$ onto pixels, $x'$, that we want wavelengths for\;
	}
\caption{Generating a Wavelength Solution}
\end{algorithm}

@@

Rather than using a simultaneous calibration fiber, two to three LFC exposures are interspersed between science exposures every ~15-30 minutes.  ThAr exposures are taken at the beginning and end of each night.  All calibration data is taken through the science fiber, thereby traveling through the same optics and hitting the same pixels as all science observations.

The results and discussion presented here are based off of 1227 LFC exposures taken between October 14 and December 18, 2020 on 29 unique nights.  LFC lines cover 40 echelle orders, which contain 19203 lines.  Though we primarily work with LFC data, we at times discuss applications to arc lamps.  These results use 78 ThAr exposures that were taken on the same nights.  ThAr lines cover all 86 extracted orders of \expres, which includes 5295 lines.

 \name\ proceeds from a given list of echelle orders, line wavelengths, and pixel positions of each line for each exposure.  We, therefore, use line positions generated by the pre-existing \expres\ pipeline, as described in \cite{petersburg2020}.

@@

 We satisfy the assumption of stability needed for \name\ by using exposures from only one ``instrumental epoch" of \expres.  \expres\ exposures are separated into different instrumental epochs that correspond to atypical hardware changes--changes that altered the position of the echellogram on the detector, the shape of the instrumental PSF, or the calibration sources.  Significant instrumental changes break the assumption of a stabilized instrument with only low-dimensional variations.  We therefore treat each instrument epoch independently.

\section{Tests}\label{sec:tests}
We perform a series of tests to validate the performance of \name\ and benchmark \name -generated wavelengths against wavelengths generated by a classic, parametric method.  For all these tests, we generate wavelengths for known calibration lines and compare the predicted wavelength to the assigned wavelength of each line.  This inherently folds in error from the measured line center of each calibration line, but the contribution to the residuals will be the same across all tests.

@@

Histograms of the residuals for each of the above described polynomial, interpolation, and denoising tests (respectively) are shown in Figure \ref{fig:testHists}.  Note that the spread in residuals is much smaller for both the denoising and interpolation tests relative to the results of the polynomial model.

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and assigned wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The standard deviation of the residuals for each model is given in the top-right corner in each model's corresponding color.}
\label{fig:testHists}
\end{figure}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/lineResids2D.png}
%\includegraphics[width=\textwidth]{Figures/lineResids2D.pdf}
\caption{Residuals plotted with respect to detector position (as defined by echelle order and x-pixel) for various wavelength models.  Each line is colored by the difference between the predicted wavelength and the true wavelength for each line, given in units of $/mps$.  Residuals to a classic, polynomial fit is shown in the left-most plot.  The center plot shows residuals to an interpolated wavelength solution.  The right-most plot shows residuals when a ThAr wavelength solution is interpolated onto LFC line positions.  The position of ThAr lines are over-plotted in black.}
\label{fig:resid2d}
\end{figure*}

@@

The residuals of the classic model is shown in the left-most plot of Figure \ref{fig:resid2d}.  There is a lot of vertical structure and some hints of a periodic, diagonal structure as well.

The residuals of the interpolation test for the same exposure is shown in the center plot of Figure \ref{fig:resid2d}.  There is no coherent structure here and smaller residuals.  We believe the flexibility of the interpolated model was able to account for high-order detector defects, which emerged as structure in the residuals of the classic, smooth, polynomial model.

@@

Though the interpolated wavelength solution returns lower, less-structured residuals than the polynomial model when guided by LFC lines, we see in the right-most plot of Figure \ref{fig:resid2d} that an interpolated wavelength solution can return much worse results in the space between widely separated calibration lines.

@@

Note that running \name\ informed by only ThAr lines cannot be regarded as a direct comparison to the LFC runs, as the increased error and variability in ThAr line positions alone makes the resultant wavelength predictions an order of magnitude worse, hence the different scale of the colorbar.  All the same, the residuals are relatively worse where lines are further spaced (for example, in redder echelle orders) than where lines are denser.

>>>>

\documentclass[12pt, onecolumn]{aastex63}
\usepackage{xcolor}
\newcommand{\lz}[1]{\textcolor{orange}{#1}}

@@

% typesetting words
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\name}{\project{Excalibur}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\expres}{\project{\acronym{EXPRES}}}
\newcommand{\eprv}{\acronym{EPRV}}
\newcommand{\lfc}{\acronym{LFC}}
\newcommand{\code}[1]{\texttt{#1}}

@@

\begin{document}
\title{\name:
A Non-Parametric, Hierarchical Wavelength-Calibration Model for a Precision Spectrograph}

@@

\begin{abstract}
\name\ is a hierarchical, non-parametric framework for wavelength calibrating precision spectrographs.  This method capitalizes on recent hardware improvements in highly-stabilized, extreme-precision radial-velocity spectrographs.  With stabilized spectrographs, the full optical system and detectors can only vary within a tiny range of configurations.  This allows us to use all calibration data to determine the entire space of possible calibrations for a stabilized spectrograph.  A wavelength solution then need only pinpoint where in this low-dimensional calibration space the spectrograph exists.  \name\ also takes advantage of the development of laser-frequency combs or etalons, which generate a dense set of stable calibration points.  This density of points permits the freedom of a non-parametric wavelength solution that can match or adapt to any instrument or detector oddities, more so than any one polynomial.  We demonstrate the success of this method with data from the EXtreme PREcision Spectrograph, \expres, which includes a Menlo Systems laser frequency comb.  We find that \name\ returns more accurate wavelength predictions by more than a factor of 2 over classic, non-hierarchical, polynomial solutions.  Employing \name\ wavelengths on \expres\ RV measurements on HD 3651 reduced the RMS of a planet fit by \lz{blah}.
\end{abstract}

% Add official keywords
\keywords{instrumentation: spectrographs -- instrumentation: detectors -- techniques: spectroscopic -- techniques: radial velocities -- methods: data analysis -- methods: statistical}

\section{Introduction} 
Extremely precise, radial-velocity (\eprv) programs have been incredibly fruitful in finding and characterizing extra-solar planets \citep{mayor2011, bonfils2013, plavchan2015, butler2017}.  These programs typically make use of spectrographs with resolutions on the order of $10^5$, which correspond to line widths on the order of $3000\,\mps$.  Exoplanet science happens at the level of $3\,\mps$ precision, with current instruments hoping to reach $0.1\,\mps$ precision \citep{fischer2016}.  This requires new spectrographs to be calibrated or stabilized to better than $10^{-4}$ of a pixel (assuming that the spectrographs are well sampled).    Some hardware--software systems, such as ESPRESSO, \expres\, and NEID were designed to deliver precision at this level \citep{pepe2013,  jurgenson2016, neid}.  Early data from \expres\ and ESPRESSO suggest that the hardware improvements do propagate through to sub-$\mps$ uncertainties on final, on-sky RV measurements \citep{blackman2020, petersburg2020, pepe2020}.

\lz{review of how wavelength solution has been done in the past.  Make the point that traditional wavelength solutions only make use of each calibration information independently
i.e. the standard method for wavelength calibration is...citations}

Here we propose to simplify and improve calibration programs for \eprv\ hardware systems with two very simple but innovative ideas.  The first flows from the observation that calibration sources---which include arc lamps (in some wavelength ranges), etalons, and laser-frequency combs (\lfc s)---illuminate the spectrograph with very stable, very dense sets of lines; almost every location in the spectrograph image plane is surrounded by nearby, useful calibration lines.

This recommends a calibration methodology that is \emph{non-parametric}:  If every point in the spectrograph detector is surrounded by nearby calibration lines, the wavelength solution can, for example, be made simply as an interpolation of the calibration data.  The density of lines removes the need to enforce any functional form for the wavelength solution (such as a continuous eighth-order polynomial, for example).  In some ways, this is a generalization of the recent work that demonstrated the efficacy of constructing a wavelength solution as multiple, segmented polynomials \citep{milakovic2020}.  Going non-parametric will improve calibration accuracy by not forcing the choice of a parametric form that may bias the calibration, especially when the chosen function is inappropriate (as, for example, polynomials will be at detector edges).

The second simple idea follows from the observation that contemporary \eprv\ instruments are incredibly stable.  Temperature-controlled, fiber-fed spectrographs vary only slightly over the night or season, and only along a small number of axes in what you might call ``calibration space'', or the (very high dimensional) space of all possible wavelength solutions.  That is, not only are the spectrographs stable, but they also have few environmentally accessible degrees of freedom.  This renders it inadvisable to fit each calibration exposure or calibrate each science exposure independently.  Instead, all the calibration data (or all the data) should be used to determine the space in which the instrument can and does vary.  Subsequent calibration work then need only determine where in the accessible part of calibration space the spectrograph was located for each exposure.

In the context of probabilistic models, this structure is \emph{hierarchical}:  The calibration data are used not just to determine the wavelength solution, but also to determine the possible \emph{space} of wavelength solutions.  In the statistics literature this concept is often described as \emph{de-noising}:  Every calibration exposure contains information about every other exposure.  Thus every exposure can be improved (i.e. de-noised) with information from every other exposure.

The method we propose here---\name---embodies these ideas.
It is a non-parametric, hierarchical, data-driven model for the wavelength solution.  By being non-parametric, it delivers enormous freedom to the wavelength solution to match or adapt to any instrument or detector oddities.  By being hierarchical, it restricts that freedom tremendously, but it does so appropriately for the empirically exercised variations in the stabilized spectrograph.  \name\ is designed for temperature-controlled, fiber-fed spectrographs with good calibration sources, such as laser-frequency combs, or etalons.  We have in mind the \eprv\ instruments and \eprv\ science cases, but we expect \name\ to have applications for other kinds of spectrographs in other contexts.  The motivating ideas behind this project are true for almost every astronomical spectrograph.

\section{Method} \label{sec:method}
\name\ is designed to turn a series of calibration lines with known wavelengths and well-fit detector positions, and de-noise and interpolate them into a full wavelength model applicable to all exposures taken with the instrument.  \name\ operates on the two core ideas that the wavelength solution should be given enormous flexibility, but that it lives in a very low-dimensional space, where the degrees of freedom are set by the limited kinematics of the stabilized spectrograph hardware.

\name\ assumes dense enough calibration line coverage with well-fit line centers to provide some constraint on an interpolated wavelength solution across an echelle order.  Additionally, \name\ assumes the instrument hardware is stabilized enough that the space of possible calibration states is low-dimensional.

Wavelength calibration is usually posed in the following way.  Given an exposure $n$, and echelle order $m$, there is a relationship between
the two-dimensional $(x,y)$-position on the detector and the
wavelength $\lambda$
\begin{equation}
\lambda(x,y,m,n) = f(x,y,m;\theta_{n})
\quad ,
\label{eq:wsol}
\end{equation}
where $\theta_{n}$ represents the parameters describing a given exposure.

Classically, pipelines employ polynomials to construct smooth wavelength solutions for each exposure.  For example, the \expres\ pipeline sets the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} to be a 2D, 9\textsuperscript{th}-order polynomial where $\theta_{n}$ represents the polynomial coefficients, $c_{nij}$, unique to each exposure $n$ \citep{petersburg2020}.
\begin{equation}
\lambda(x,m,n) = \sum_{i=0}^9\sum_{j=0}^9 c_{nij}\, x^i\,m^j + \mathrm{noise}
\quad ,
\label{eq:poly_wsol}
\end{equation}
where the y-dependence is dropped as that dependence is carried by spectral order $m$.  The coefficients $c_{nij}$ are interpolated to time $t_n$ of exposure $n$ by a third-order polynomial of time.  This third-order polynomial is evaluated at the time of non-calibration exposures to re-construct a 2D, 9th-order polynomial wavelength solution for that exposure.  Each calibration image obtains in $c_{nij}$ coefficients independently.

@@

If the space of all calibration possibilities is in fact $K$-dimensional (where $K$ is a small integer, i.e 2 or 8 or thereabouts), and if the calibration variations are so small that than can be linearized, then the function $f(x,m;\theta_{n})$ from Equation \ref{eq:wsol} should be low-dimensional.  In \name, we transpose the calibration model---making the position $x$ a function of $\lambda$---into the following form
\begin{equation}
x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)
\quad ,
\label{eq:excl_wsol}
\end{equation}
where
$g_0(\lambda,m)$ is the fiducial or mean or standard calibration of the spectrograph,
the $a_{nk}$ are $K$ scalar amplitudes for each exposure $n$,
and the $g_k(\lambda,m)$ are basis functions expressing the ``directions'' in calibration space that the spectrograph can depart from the fiducial calibration.  When this calibration structure is used to deliver a wavelength solution, we will have to invert $x(\lambda,m,n)$ into $\lambda(x,m,n)$.

The challenge is to learn these basis functions from the data and get the $K$ amplitudes, $a_{nk}$, for every exposure $n$.  There are many ways to discern these basis functions.  In this paper, we present a model using principal component analysis (PCA).  A PCA is justifiable in the limit where there is very high signal-to-noise ratio, as is usually the case with typical calibration images.  There are many alternatives to PCA for this dimensionality reduction; we return to this point in the discussion (\textsection \ref{sec:discussion}) below.

\subsection{Dimensionality Reduction: De-Noising of Calibration Frames} \label{sec:denoising}
Within a given range of stability, \name\ will use calibration images to  1) determine the space in which the instrument varies and 2) where in the accessible calibration space the spectrograph existed for each exposure.  For each calibration exposure, $n$, \name\ requires a full list of lines, $(\lambda,m)$ that are expected in each calibration exposure.  Each line is uniquely defined by a combination of echelle order, $m$, and ``true'' or theoretical wavelength, $\lambda$.

For each exposure, $n$, every line, $(\lambda,m)$, has an associated fitted detector position, $x(\lambda,m,n)$, for example x-pixel in an 2D extracted echelle order.  Fitted line centers that are missing from an exposure (e.g. because the fit failed due to noise, the line is not in its usual echelle order, etc.) can be assigned a \code{NaN} (hardware not-a-number) for that echelle order instead.  Let there be $M$ lines per exposure.  \name\ reads in a $N \times M$ matrix of line positions for each of $M$ lines for each of $N$ exposures.

The mean of measured line position over the set of calibration exposures represents the fiducial, or standard calibration of the spectrograph, $g_0(\lambda,m)$.  In this implementation of \name\, principal component analysis is performed on the difference between this fiducial calibration and each individual line position.  The returned principal components serve as basis functions,  $g_k(\lambda,m)$, expressing the possible deviations of the spectrograph from this fiducial calibration.  The magnitude of each principal component for each exposure, $a_{nk}$, represents the scalar amplitude of these deviations for each exposure.  \name\ then uses a small number, $K$, of principal components to reconstruct a de-noised version of the line positions as formulated in Equation \ref{eq:excl_wsol}.

Missing line center measurements are replaced with de-noised estimates.  This is done iteratively until the estimates of missing line centers change by less than 0.01\%.  This process can be repeated on line centers deemed as outliers by some metric, to account for lines that may have been mis-identified or mis-fit.  The principal components from the final iteration are used to define the spectrograph's calibration space while the associated magnitudes pinpoint where in that calibration space the spectrograph is located for each calibration exposure.

\begin{algorithm}
\SetAlgoLined
\KwData{line positions $x(\lambda,m,n)$ for each exposure $n$, with wavelengths $\lambda$ and echelle orders $m$}
\KwResult{Low-dimensional calibration space of spectrograph}
\While{change in missing or outlier line centers $>$ 0.01\%}
{
	$g_0(\lambda,m) = \overline{x(\lambda,m,n)}$\;
	find $U, \Sigma, V$ s.t. $U\Sigma V^* = (x(\lambda,m,n)-g_0(\lambda,m))$\;
	let $a_{n,k} = U\cdot \Sigma$ and $g_k(\lambda,m) = V$\;
	$x(\lambda,m,n) = g_0(\lambda,m) + \sum_{k=1}^K a_{nk}\,g_k(\lambda,m)$ where $K=6$ for $x(\lambda,m,n)$ = \code{NaN}
	}
\caption{Dimension Reduction and De-Noising}
\end{algorithm}

\subsection{?Interpolating Calibration Frames?}
 \label{sec:interp_time}
The magnitude of each principal component can be interpolated with respect to housekeeping data, such as time, temperature, telescope position, etc. to determine different calibration states of the spectrograph.  The choice of what to interpolate against depends on the dominant contribution to variation in the instrument.

In the implementation of \name\ presented here, the magnitudes of the principal components are interpolated with respect to time.  New magnitudes $a_n'k$ values are found for the midpoint time of non-calibration exposures and used to construct the calibration state of the spectrograph for that exposure.  Using interpolated magnitudes, $a_{n;k}$ and the basis vectors, $g_k(\lambda,m)$, returned by the de-noising process, \name\ can construct a set of calibration lines, $x(\lambda,m,n') $, for any exposure as prescribed in Equation \ref{eq:excl_wsol}.

\subsection{Interpolating a Wavelength Solution} \label{sec:interp_wsol}
From the de-noising step, \name\ can now construct a set of calibration lines, $x(\lambda,m,n')$ for any exposure $n'$.  To construct a wavelength solution, \name\ then interpolates the known wavelengths over the exposure specific line centers.  For instance, interpolating the known wavelengths vs. line centers onto every integer x will generate wavelengths for each pixel in an echelle order.

\begin{algorithm}
\SetAlgoLined
\KwData{the fiducial calibration of the spectrograph $g_0(\lambda,m)$; magnitudes of the principal components for each exposure $a_{n,k}$; basis vectors spanning the calibration space of the spectrograph $g_k(\lambda,m)$; }
\KwResult{Wavelengths for detector positions $x'(m,n')$ of exposure $n'$ with time $t_{n'}$}

Find $a_{n',k}$ by interpolating $a_{n,k}$ with to $t_{n'}$\;
$x(\lambda,m,n') = g_0(\lambda,m) + \sum_{k=1}^K a_{n'k}\,g_k(\lambda,m)$ where $K=6$\;
\For{each unique $m$}{
	interpolate $\lambda$ with respect to $x(\lambda,m,n')$ onto pixels $x'(m,n')$\;
	}
\caption{Generating Wavelength Solution}
\end{algorithm}

@@

Rather than using a simultaneous calibration fiber, two to three LFC exposures are interspersed between science exposures every 15-30 minutes.  ThAr exposures are taken at the beginning and end of each night.  All calibration data is taken through the science fiber, meaning they traveling down the same optical pathway and are projected onto the same pixels as all science observations.

The results and discussion presented here are based off of 1227 LFC exposures taken between October 14 and December 18, 2020 on 29 unique nights.  LFC lines cover 40 echelle orders, which contain 19203 lines.  Though the results are primarily based on work with LFC data, there is some discussion of applications to arc lamps.  These results use 78 ThAr exposures that were taken on the same nights.  ThAr lines cover all 86 extracted orders of \expres, which includes 5295 lines.  For both the LFC and ThAr data, lines that appeared in less than 60\% of exposures were not included in the analysis.  Similarly, files with more than 60\% of expected lines missing were cut from the analysis.

 \name\ proceeds from a given list of echelle orders, line wavelengths, and pixel positions of each line for each exposure.  We, therefore, use line positions generated by the pre-existing \expres\ pipeline \citep{petersburg2020}.

@@

In order to satisfy the assumption of stability needed for \name, only exposures from only one ``instrumental epoch" of \expres\ are used.  \expres\ exposures are separated into different instrumental epochs that correspond to atypical hardware changes--changes that altered the position of the echellogram on the detector, the shape of the instrumental PSF, or the calibration sources.  Significant instrumental changes break the assumption of a stabilized instrument with only low-dimensional variations.  Each instrument epoch is therefore treated independently.

\section{Tests}\label{sec:tests}
We perform a series of tests to validate the performance of \name\ and benchmark \name -generated wavelengths against wavelengths generated by a classic, parametric method.  For all these tests, we generate wavelengths for known calibration lines and compare the predicted wavelength to the assigned wavelength of each line.  This inherently folds in errors in the measured line center of each calibration line, but the contribution to the residuals will be the same across all tests.

@@

Histograms of the residuals for each of the above described polynomial, interpolation, and denoising tests (respectively) are shown in Figure \ref{fig:testHists}.  Note that the spread in residuals is much smaller for both the denoising and interpolation tests relative to the results of the polynomial wavelength solutionl.

\begin{figure}[h]
\centering
\includegraphics[width=.5\textwidth]{Figures/all_results.pdf}
\caption{Difference in predicted and assigned wavelengths for the wavelength calibration tests described in Section \ref{sec:tests}.  The standard deviation of the residuals for each test is given in the top-right corner in each method's corresponding color.}
\label{fig:testHists}
\end{figure}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{Figures/lineResids2D.png}
%\includegraphics[width=\textwidth]{Figures/lineResids2D.pdf}
\caption{Residuals plotted with respect to detector position (as defined by echelle order and x-pixel) for various wavelength calibration methods.  Each line is colored by the difference between the predicted wavelength and the true wavelength for each line, given in units of $/mps$.  Residuals to a classic, polynomial fit is shown in the left-most plot.  The center plot shows residuals to an interpolated wavelength solution.  The right-most plot shows residuals when a ThAr wavelength solution is interpolated onto LFC line positions.  The position of ThAr lines are over-plotted in black.}
\label{fig:resid2d}
\end{figure*}

@@

The residuals of the classic, polynomial wavelength solution is shown in the left-most plot of Figure \ref{fig:resid2d}.  There is a lot of vertical structure and some hints of a periodic, diagonal structure as well.

The residuals of the interpolation test for the same exposure is shown in the center plot of Figure \ref{fig:resid2d}.  There is no coherent structure here and smaller residuals.  We believe the flexibility of the interpolated model was able to account for high-order detector defects, which emerged as structure in the residuals of the classic, smooth, polynomial-driven wavelength solution.

@@

Though the interpolated wavelength solution returns lower, less-structured residuals than the polynomial wavelength solution when guided by LFC lines, we see in the right-most plot of Figure \ref{fig:resid2d} that an interpolated wavelength solution can return much worse results in the space between widely separated calibration lines.

@@

Note that running \name\ informed by only ThAr lines cannot be regarded as a direct comparison to the LFC runs, as the increased uncertainty and variability in ThAr line positions alone makes the resultant wavelength predictions an order of magnitude worse, hence the different scale of the colorbar.  All the same, the residuals are relatively worse where lines are further spaced (for example, in redder echelle orders) than where lines are denser.
